Sending a Balloon to the Edge of Space

### 2018-11-14 paper/paper.rmd
almost done with paper

---
title: "Sending a Balloon to the Edge of Space"
author: Luker Bowsher, John Kim, Vivian Liu, Simon Oros, Caroline Pang, Alec Vercruysse
date: "November 13, 2018"
output: 
  pdf_document:
    number_sections: true
    keep_tex: true
    includes:
      in_header: "preamble.tex"
fontsize: "12pt"
indent: true
abstract: This paper details 6thsense, a mission to send a weather balloon to to edge of space in order to study Earth's atmosphere. The weather balloon is outfitted with a payload that includes a camera, inside and outside temperature sensors, a barometer, multiple GPS tracking devices, as well as a host of individual experiments. The balloon most likely reached an altitude of 20,000 meters, into the stratosphere and ozone layer. The balloon was successfully recovered after approximately one and a half hours of flight.
---

\newpage
\setcounter{secnumdepth}{2}
\setcounter{tocdepth}{2}
\tableofcontents
\newpage
\onehalfspacing

```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
library(gridExtra)
library(latex2exp)
opts_chunk$set(echo = F, warning = F)

@@

There are four main layers to the atmosphere: the troposphere, stratosphere, mesosphere, and thermosphere. The **troposphere** is the section of the atmosphere from roughly 0km to 15km, and is the layer in which we live. It includes 75% of the mass of all gases in the atmosphere, consisting of 78% nitrogen gas, 21% oxygen, and less than 1% of argon, carbon dioxide, and other constituents. Because it is both the lowest in altitude and most massive layer, the pressure in the troposphere is the greatest: 100kPa. As altitude increases in the troposphere, temperature decreases to about -60C. The** stratosphere** is the section from roughly 15km to 50km. The ozone layer, or the region in the atmosphere which absorbs most UV rays, is within the stratosphere at about 20km in altitude. The pressure in the stratosphere is 10kPa. The mesosphere is the region from 50km to 80km. The pressure in the mesosphere is 0.1 kPa. Finally, the thermosphere is the highest in altitude of the four layers, spanning from 80km into outer space. Here, the pressure is only 0.01kPa, as it is the least dense of the layers. The few air molecules in the thermosphere are heated to extremely high temperatures by the Sun’s radiation. However, because there are so few molecules, we cannot measure an appreciable difference in overall temperature, so the effective temperature of the thermosphere is less than that of the mesosphere.

The layers relevant to our experiment are the troposphere and stratosphere, as our balloon launch was designed to reach a maximum altitude of 30km (about the middle of the stratosphere but past the ozone layer). Therefore, we had to prepare the payload for temperatures as low as -60C and pressure as low as 10kPa (in the stratosphere). We expect to encounter the lowest temperature in the troposphere, around an altitude of 10km. As the balloon rises from sea level to the top of the troposphere, we expect to see temperature decrease because the ground absorbs most of the sun’s radiation (as opposed to the gas particles). However, as the balloon enters the stratosphere, we expect an increase in temperature due to the ozone layer’s blockage of UV rays. The UV rays that the ozone layer absorbs causes the gas particles in the stratosphere to heat up, so temperature increases with altitude. As the balloon gains altitude, the density of air around it decreases, so the pressure decreases.

@@

When infrared radiation enters the atmosphere, short-wave infrared radiation (700 -  1000 nm) does not effectively interact with greenhouse gasses. The photons instead pass through the atmosphere and reach the surface of the earth. Here, they are absorbed by the surface of the earth and get reemitted as a lower energy, longer-waved infrared photon. Greenhouse gasses can now absorb these lower energy photons and reemit them. This absorption and reemission warms the troposphere by increasing the kinetic energy the atmosphere. Additionally, greenhouse gasses can reemit photons back towards the surface of the earth, increasing the temperature of the earth’s surface. By impeding the tendency for long-wave radiation to leave the atmosphere, the greenhouse effect increase the temperature of the troposphere and surface.

The spectroscopy that explains the greenhouse effect concerns molecular vibration. Infrared photons excite vibrations in molecules. Since the vibrational frequency of atoms are quantized, only certain energies will excite vibration in a molecule. Different molecules can exhibit different vibrational modes. A molecule with more atoms has more vibrational modes. The infrared selection rule states that a vibrational mode is active when it is changing the dipole moment of the molecule. An active vibrational mode is a molecular vibration that can be excited by an infrared photon. Thus, infrared photons will only be absorbed when that molecule’s dipole moment is changing. Molecules such as nitrogen and oxygen will never exhibit a change in their dipole moment due to vibration because their only vibrational mode involves the singular oscillation of the distance between two atoms. Thus, they cannot absorb infrared radiation in order to partake in the greenhouse effect and are not considered greenhouse gasses. R On the other hand, many atmospheric molecules that have more than two atoms have more complex vibrational modes. These active vibrational mode gasses, such as nitrogen dioxide and water, are called greenhouse gasses.

As the greenhouse effect warms the planet, many factors accelerate the temperature rise by creating positive feedback loops. Positive feedback loops amplify any global increase in temperature. One positive feedback loops is created when rising temperatures melt glaciers. These white and reflective icy surfaces melt into dark oceans which absorb more radiation and heat up faster. This explains the approaching transition from a permanent to a seasonal ice cover in the Arctic Ocean. Thus temperature increases accelerate.^[https://www.nature.com/articles/s41598-017-08467-z] Another positive feedback loops involves the melting of permafrost. A quarter of the northern hemisphere is covered in permafrost that holds concentrated methane and carbon dioxide. Slight growths in temperature are enough to thaw the permafrost. When it melts, these greenhouse gasses are released into the atmosphere. Thus, the greenhouse effect amplifies and global temperature growth accelerates. ^[http://science.sciencemag.org/content/340/6129/183] Finally, humidity is also a positive feedback loop. As global temperatures rise, the atmosphere can hold more water vapor because less of it gets condensed. Water vapor is a greenhouse gas itself. Thus any increase in global temperatures will be amplified due to the water vapor positive feedback loop.

The annual increase of atmospheric carbon dioxide concentration is approximately 100 times greater in the past 60 years than it has been during any previous natural increase.F Additionally, methane concentrations have more than doubled since the start of the industrial revolution.B  Partly due to the increase of these gas concentration along with many positive feedback loops, the planet's average surface temperature has risen 0.9\degree C since the late 19th century.F With rising temperatures comes the consequences of severe weather patterns, flooding and other natural disasters. Collecting atmospheric data of things such as temperature and gas concentrations informs our understanding of the progression of the greenhouse effect. With this information, we can begin to look at ways to impede rising global temperatures.

The Panel on Climate Change has recognized that as global temperatures rise, lightning strikes will occur more frequently. As our skyscrapers get higher and multiply, lightning strike sourced Terrestrial Gamma ray Flashes (TGF) become a concern.

@@

While there is an abundance of research done on upward directed TGFs (those detected by the satellites in the 1950s), there is a lack of data detecting downward directed TGFs. Currently, researchers led by David M. Smith at UC Santa Cruz and other researchers around the world are using a multitude of methods to detect downwards directed TGFs. Some methods include sending detectors in cargo planes, on weather balloons and on the ground in thunder prone areas. Since TGFs are so powerful, detectors usually fail if they are near a TGF because of the sheer amount of energy exerted. This field of research is relatively new and active.

**INSERT PRESSURE MODEL ANALYSIS HERE**

The exponential model for temperature vs altitude, where altitude is less than 11km (or where altitude is within the troposphere) is: T=T0(1-h/44329), where h=height in meters, and T0=temperature at sea level. This relationship shows a negative relationship between height and temperature. As altitude increases within the troposphere, temperature decreases.

@@

@@

**INCLUDE INDIVIDUAL MOTIVATIONS HERE**

@@

lorem ipsum

@@

lorem ipsum

@@

lorem ipsum

# Tracking The Payload

>>>>

---
title: "Sending a Balloon to the Edge of Space"
author: Luker Bowsher, John Kim, Vivian Liu, Simon Oros, Caroline Pang, Alec Vercruysse
date: "November 13, 2018"
output: 
  pdf_document:
    number_sections: true
    keep_tex: yes
    includes:
      in_header: "preamble.tex"
    fig_caption: yes
fontsize: "12pt"
indent: true
abstract: This paper details 6thsense, a mission to send a weather balloon to to edge of space in order to study Earth's atmosphere. The weather balloon is outfitted with a payload that includes a camera, inside and outside temperature sensors, a barometer, multiple GPS tracking devices, as well as a host of individual experiments. The balloon most likely reached an altitude of 20,000 meters, into the stratosphere and ozone layer. The balloon was successfully recovered after approximately one and one half hours of flight.
---

\newpage
\setcounter{secnumdepth}{3}
\setcounter{tocdepth}{3}
\tableofcontents
\newpage
\onehalfspacing

```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
library(gridExtra)
library(latex2exp)
opts_chunk$set(echo = F, warning = F, fig.align='center', fig.pos = 'h')

@@

## Background

There are four main layers to the atmosphere: the troposphere, stratosphere, mesosphere, and thermosphere. The **troposphere** is the section of the atmosphere from roughly 0km to 15km, and is the layer in which humans live. It includes 75% of the mass of all gases in the atmosphere, consisting of 78% nitrogen gas, 21% oxygen, and less than 1% of argon, carbon dioxide, and other constituents. Because it is both the lowest in altitude and most massive layer, the pressure in the troposphere is the greatest: 100kPa. As altitude increases in the troposphere, temperature decreases to about -60 \degree C. The** stratosphere** is the section from roughly 15km to 50km. The ozone layer, or the region in the atmosphere which absorbs most UV rays, is within the stratosphere at about 20km in altitude. The pressure in the stratosphere is 10kPa. The mesosphere is the region from 50km to 80km. The pressure in the mesosphere is 0.1 kPa. Finally, the thermosphere is the highest in altitude of the four layers, spanning from 80km into outer space. Here, the pressure is only 0.01kPa, as it is the least dense of the layers. The few air molecules in the thermosphere are heated to extremely high temperatures by the Sun’s radiation. However, because there are so few molecules, an appreciable difference in overall temperature cannot be measured, so the effective temperature of the thermosphere is less than that of the mesosphere.

The layers relevant to the experiments conducted are the troposphere and stratosphere, as the balloon was designed to reach a maximum altitude of 30km (about the middle of the stratosphere but past the ozone layer). Therefore, the payload had to prepared for temperatures as low as -60 \degree C and pressure as low as 10kPa (in the stratosphere). The payload is expected to encounter the lowest temperature in the troposphere, around an altitude of 10km. As the balloon rises from sea level to the top of the troposphere, temperature is expected to decrease because the ground absorbs most of the sun’s radiation (as opposed to the gas particles). However, as the balloon enters the stratosphere, an increase in temperature is expected due to the ozone layer’s blockage of UV rays. The UV rays that the ozone layer absorbs causes the gas particles in the stratosphere to heat up, so temperature increases with altitude. As the balloon gains altitude, the density of air around it decreases, so the pressure decreases.

## Greenhouse Gases in the Atmosphere

@@

When infrared radiation enters the atmosphere, short-wave infrared radiation (700 -  1000 nm) does not effectively interact with greenhouse gasses. The photons instead pass through the atmosphere and reach the surface of the earth. Here, they are absorbed by the surface of the earth and get reemitted as a lower energy, longer-waved infrared photon. Greenhouse gasses can now absorb these lower energy photons and reemit them. This absorption and reemission warms the troposphere by increasing the kinetic energy the atmosphere. Additionally, greenhouse gasses can reemit photons back towards the surface of the earth, increasing the temperature of the earth’s surface. By impeding the tendency for long-wave radiation to leave the atmosphere, the greenhouse effect increase the temperature of the troposphere and surface. (Britannica, 2018)

The spectroscopy that explains the greenhouse effect concerns molecular vibration. Infrared photons excite vibrations in molecules. Since the vibrational frequency of atoms are quantized, only certain energies will excite vibration in a molecule. Different molecules can exhibit different vibrational modes. A molecule with more atoms has more vibrational modes. The infrared selection rule states that a vibrational mode is active when it is changing the dipole moment of the molecule. An active vibrational mode is a molecular vibration that can be excited by an infrared photon. Thus, infrared photons will only be absorbed when that molecule’s dipole moment is changing. (Pecsok, 1968) Molecules such as nitrogen and oxygen will never exhibit a change in their dipole moment due to vibration because their only vibrational mode involves the singular oscillation of the distance between two atoms. Thus, they cannot absorb infrared radiation in order to partake in the greenhouse effect and are not considered greenhouse gasses. On the other hand, many atmospheric molecules that have more than two atoms have more complex vibrational modes. These active vibrational mode gasses, such as nitrogen dioxide and water, are called greenhouse gasses. (UCAR, 2012)

As the greenhouse effect warms the planet, many factors accelerate the temperature rise by creating positive feedback loops. Positive feedback loops amplify any global increase in temperature. One positive feedback loop is created when rising temperatures melt glaciers. These white and reflective icy surfaces melt into dark oceans which absorb more radiation and heat up faster. This explains the approaching transition from a permanent to a seasonal ice cover in the Arctic Ocean. Thus temperature increases accelerate. (Kashiwase, 2017)

Another positive feedback loop involves the melting of permafrost. A quarter of the northern hemisphere is covered in permafrost that holds concentrated methane and carbon dioxide. Slight growths in temperature are enough to thaw the permafrost. When it melts, these greenhouse gasses are released into the atmosphere. Thus, the greenhouse effect amplifies and global temperature growth accelerates. (Vaks, 2013)

Finally, humidity is also a positive feedback loop. As global temperatures rise, the atmosphere can hold more water vapor because less of it gets condensed. Water vapor is a greenhouse gas itself. Thus any increase in global temperatures will be amplified due to the water vapor positive feedback loop.

The annual increase of atmospheric carbon dioxide concentration is approximately 100 times greater in the past 60 years than it has been during any previous natural increase (Lindsey, 2018). Additionally, methane concentrations have more than doubled since the start of the industrial revolution (Britannica, 2018).  Partly due to the increase of these gas concentration along with many positive feedback loops, the planet's average surface temperature has risen 0.9\degree C since the late 19th century (Lindsey). With rising temperatures comes the consequences of severe weather patterns, flooding and other natural disasters. Collecting atmospheric data of things such as temperature and gas concentrations informs our understanding of the progression of the greenhouse effect. With this information, methods can be researched and developed to impede rising global temperatures.

The Panel on Climate Change has recognized that as global temperatures rise, lightning strikes will occur more frequently. As skyscrapers get higher and multiply, lightning strike sourced Terrestrial Gamma ray Flashes (TGF) become a concern.

@@

While there is an abundance of research done on upward directed TGFs (those detected by the satellites in the 1950s), there is a lack of data detecting downward directed TGFs. Currently, researchers led by David M. Smith at UC Santa Cruz and other researchers around the world are using a multitude of methods to detect downwards directed TGFs. Some methods include sending detectors in cargo planes, on weather balloons and on the ground in thunder prone areas. Since TGFs are so powerful, detectors usually fail if they are near a TGF because of the sheer amount of energy exerted. This field of research is relatively new and active. (Smith, 2015)

##Empirical and Exponential Models for Temperature and Pressure

```{r NASA_image, fig.cap="\\label{fig:figs} NASA Empirical Models for Temperatur and Pressure"}
include_graphics("assets/nasa_model.jpeg")
```

Figure 1 shows NASA’s empirical models for pressure and temperature for the troposphere and stratosphere. For a height of 0, the temperature is that of sea level, or 15.06 \degree C. At a height of 11km, the predicted temperature is -56.35 \degree C, which is consistent with measurements recorded by the payload. As the balloon rises through the lower stratosphere, a constant temperature of -56.46 \degree C is expected. Finally, once the balloon rises beyond the ozone layer (around 25km altitude), temperatures are expected to rise again due to the ozone layer’s absorption of the sun’s radiation. Refer to section 8 for an overlay of NASA’s empirical models over experimental data.

@@

### How a Pressure Sensor Works

Most pressure sensors, including the Vernier gas pressure sensor that is connected to the main Arduino mega board and used on the payload, measure gas pressure by using a flexible diaphragm, where one side is a vacuum and the other side is open to the surrounding atmosphere, and producing a voltage that varies linearly with absolute pressure. A typical capacitive humidity sensor measures relative humidity by placing a sheet of metal oxide between two electrodes and measuring the change in capacitance between the electrodes, which is directly proportional to the change in air moisture. And lastly, the temperature sensor used in the following experiments, specifically a thermistor, measures temperature through nonlinear changes in resistance, which decreases as temperature increases.

@@

### Alec Vercruysse: Spectrometer

Identifying the gaseous composition of the atmosphere is of great importance in order to be able to properly analyze the different layers and their properties. Specifically, identifying the composition of the ozone layer is extremely important in order to determine its health. Actively monitoring the ozone layer is crucial, since it is responsible for shielding life on earth from harmful Ultraviolet light. Furthermore, analyzing the gaseous composition of the troposphere is important to understanding and measuring the greenhouse effect (see section 1). The more greenhouse gasses in the atmosphere, the larger the greenhouse effect. This leads to more heat trapped in the atmosphere, and accelerated global warming.

A spectrometer was placed at a 45 \degree angle relative to the top acrylic plate in order to perform spectral analysis of the atmosphere, with a view unobstructed by the balloon or payload.

### John Kim: Methane Sensor

Greenhouse gases are gases that trap heat that come from both the human activity and the Sun inside the Earth’s atmosphere, resulting in global warming and climate change when overlooked. While many identify greenhouse gases with carbon dioxide, methane contributes to over 10% of the gaseous composition of the atmosphere and is emitted from coal mining and organic waste. Monitoring methane levels is crucial to not only furthering the understanding of how the Earth’s atmospheric composition changes over time, but also protecting the environment—methane is much more efficient than carbon dioxide at absorbing heat and more potent. Scientists have even written letters to the United Nations’ climate change summit to require nations to curb methane emissions more quickly.

A methane sensor was placed inside the insulated payload because it was found to be fairly sensitive to temperature when compared to other components that were placed outside the styrofoam box.  In addition, there would not have been any significant difference in methane levels inside and outside the payload because the non-restrictive airflow inside the box would have allowed methane gas to pass through relatively unobstructed. A typical methane sensor measures methane gas by detecting the changes in resistance when the gas comes into contact with a catalytic surface and heats the wiring resistance.

### Caroline Pang: UVB Sensor

Ultraviolet light is a harmful form of radiation which enters our atmosphere. There are three classifications of ultraviolet radiation, UVA, UVB, and UVC. The ozone layer is responsible for absorbing most of the harmful UVC radiation making it an essential part of the atmosphere. However, it only partially absorbs UVB light which is responsible for causing sunburns and in extreme cases, skin cancer. Measuring the intensity of UVB light as a function of altitude and observing how it changes when entering the ozone layer can demonstrate how effectively the ozone layer absorbs UVB light.

Therefore, a UVB light sensor was placed vertically inside the payload with the top half of the rod outside of the acrylic top to measure the intensity of UVB light.

### Vivian Liu: Light Intensity and Solar Power
Light intensity, specifically for the purposes of solar energy, is an important part of the initiative for cleaner energy. It is also interesting to learn about how light intensity increases or decreases as it passes through the ozone layer.

To test light intensity, a photoresistor was placed on a board, and the resistances of the photoresistor attached in series to a 1.2k Ohm resistor were recorded as a function of altitude. From the resistance of the photoresistor, lux, or the units of light intensity, can be calculated. For solar power, the arduino read the voltage of a solar panel that was placed outside of the shadow of the balloon. From this voltage, power output of the solar panel can be calculated. The purpose of the experiments on solar properties is to see if there is a correlation between altitude and solar output.

@@

##Payload Design

###Structure

The payload design consists of a styrofoam box and an acrylic rectangle as a lid. Styrofoam was used because it is both light and insulating. The acrylic lid was heavy, but it allowed the interior of the payload to be seen at all times. Most importantly, it provided for firm mounting points for the arduino, all circuit boards, and the sticky GoPro mount. The lid has two rectangles cut out in the middle so that wires and sensors can go from the outside to the inside of the payload. The lid has two holes along its width to accommodate two vertical wooden dowels that fasten the lid to the styrofoam box. A hole was drilled through the wall of the styrofoam box so that the dowel can be inserted and secured using epoxy glue. The wooden dowels are then doubly secured by fastening it to the box using cotter pins and washers. The payload can be connected to the balloon using the two wooden dowels that run horizontally through the top of the payload. Strings are tied to the ends of the wooden dowels, and secured using a combination of epoxy glue and cotter pins/washers. Finally, the cutdown mechanism was secured by wrapping wire around the string connecting the payload to the carabiner.

###Organization of Parts

Most of the important electronics were attached to the acrylic lid of the payload. The arduino mega and protoboard containing individual experiments are screwed onto the interior of the acrylic lid using nylon spacers and screws. These components are on the interior of the payload for protection. The BME sensor, protoboard (containing sensors), GoPro mount, and transmitter are screwed onto the exterior of the acrylic lid. The spectrometer is outside the payload to give it access to the sky, and the transmitter is also on the exterior to provide better line-of-sight. These essential electronics are secured onto the lid so as to reduce movement within the payload. Other plastic items, such as the UV sensor, gas pressure sensor, and fox-hunt transmitter, were placed within the styrofoam box. Finally, the power source, four 9V batteries, and the picoAPRS were placed at the bottom of the payload underneath a layer of fiberglass insulation. This was to help preserve thermal energy inside the payload so that the batteries could continue working even in very cold conditions, including those of the PicoAPRS.

Near the bottom of the payload, there are two 3’ carbon rods that jut out of the payload, and a wooden box at the end of the rods. This is so that the solar panel can be secured on top of the wood box, and the distance from the balloon to the end of the rods can ensure that the solar panel is not obscured by the shadow of the balloon.

####Table 1: Weights of Items in Payload
Item | Weight (g)
------------------------------------------------------------------ | -------------------
Cutdown mechanism (protoboard + 6 capacitors + wiring) | 64
PicoAPRS | 59
Fox-Hunt Transmitter | 77
4 batteries (40g/battery) | 160
**Payload cover** (arduino mega, arduino uno, two protoboards, transmitter, pressure sensor, GoPro mount, ME, UV sensor, acrylic cover, wiring) | 604
**Payload body** (styrofoam, two 3’ carbon rods, solar panel, small wood box, four 10” wood dowels) | 175
Fiberglass insulation | 30
GoPro | 200
**Total** | 1.369 kg

## The Tethered Launch Reel-in System

### Structure

The structure of the reel in system consists of a sturdy, slightly elevated base and two sides to hold the axle. The base is made from a square frame with one additional plank for the motor to be mounted on. The axle holders are each made from one vertical 2x2 post with a hole through the top to fit a bearing, and two slanted supports to form a triangular shape. In addition to this, there are two thin wooden pieces screwed on the outside of each post to prevent the axle from popping out of the frame. All of the pieces are secured with construction screws.

### Gears and Axle

The reel in system has two gears, a 72 tooth gear on the axle, and a 10 tooth gear mounted on the motor. Therefore, there is a theoretical mechanical advantage of 7.2. However, the measured mechanical advantage was around 5.9. This mechanical advantage is necessary so that the motor can provide enough torque to reel in the balloon and so that the axle does not spin too quickly.

$$MA = \frac{f_{out}}{f_{in}} = \frac{d_{in}}{d_{out}} = \frac{72}{10}$$

Securing the larger gear to the axle is one of the most important parts of the reel in system. Therefore, there are two mechanisms in place to keep the gear in place. First there are two small disks which fit snugly onto the axle on each side of the gear which are screwed together, somewhat clamping the gear to the axle. However, the disks are still able to rotate and slide along the axle when force is applied. Thus, the addition of a more secure mechanism was implemented by attaching a wooden block to the axle, drilling a hole through the block and axle, and then securing them together by a screw. The two small disks were then attached to the block by screws.

In addition to this, the axle includes two larger wooden disks spaced apart to set the boundaries for where the string is wrapped. One failure point in the reel in system was that the string was wrapped unevenly around the axle, causing it to tangle multiple times during the tethered launch. In the future, the string should be moved side to side more often when being wrapped.

### The Tensioning System

Maintaining tension in the chain is essential for the reel in system to work effectively because increases the contact between the chain and the gears, and it protects against the chain falling off the gears at high speeds. To implement a tensioning system, a thin wooden bar was first installed across axle holders to mount the system across from the chain. The system is comprised of an S-shaped piece of wood with a hole going horizontally through the top ends where an axle was placed to hold a sprocket. The bottom end of the wood was secured to the frame with a piece of string acting as a pivot point. To apply a downward force on the piece of wood, a compression spring was placed between the wood and the crossbar and held in place by screws. In operation, the spring pushed the gear onto the chain, creating tension in the chain.

## Fox-Hunting Yagi Antenna

The 3 element yagi antenna used for Fox-Hunting is built from PVC piping, cross connectors, a tape measure, hose clamps, a piece of 14 AWG wire, and a coax cable.

All three elements, the were each measured and cut to the appropriate length from the tape measure so that the antenna could receive the 147.065mHz signal from the Fox-Hunting transmitter. The reflector was cut to 41.375 in, the driven element was comprised of two 17.75 in segments, and the director was 35.125 in long. Additionally, the spacing between each element was carefully measured when cutting the PVC pipes to the correct length, with 12.5 in between the director and driven element, and 8 in between the driven element and the reflector. However, even though our Fox-Hunting was successful, the lengths and distances of each element could have been fine-tuned to increase the sensitivity of the antenna to the frequency of our Fox-Hunting transmitter. See section 6.5.

A coax cable had one end separated into the grounded shield and signal conductor, each which was attached to an opposite end of the driven element. A wire was also connected to the driven element to complete the dipole. The other end of the coax cable included a BNC connector that was attached to a radio receiver.

All physical CAD drawings are included in the appendix.

@@

```{r cutdown_diagram, fig.cap="\\label{fig:figs} Cutdown mechanism circuit design"}
include_graphics("assets/cutdown.png")
```

## Cutdown Method

The cutdown mechanism functions by  running electricity through nichrome wire in order to burn the string that attaches the payload to the balloon. Two 5 cm lengths of 26g nichrome wire were used. 5 loops of the double stranded nichrome wire were wrapped around the the string that is to be cut. $R_{2}$ represents the nichrome wire.

## Charge Needed

A stand with a 1.4 kg weight attached (the expected maximum weight of the payload) was used as a testing apparatus. First, a power supply was used to test the cutdown mechanism.  The apparatus was placed in a -47\degree C freezer for 30 minutes before testing. We chose this temperature because it is colder than the expected minimum temperature that the payload will experience. After 30 minutes, we observed that it took 6.6A for 0.71s in order to cut the string. We calculated the amount of charge needed to cut the string.

$$I=\frac{\Delta Q}{\Delta t}$$

The amount of charge needed is 4.7C.

## Capacitor Bank

While this is the minimum amount of charge required for the cutdown to work, significantly more charge was stored in order to combat leaky capacitors and cold temperatures. Our capacitor bank is composed of supercapacitors ($C_{1}-C_{6}$). Each capacitor stores Three capacitors were wired up in series. This increased the voltage of the capacitor bank threefold. We chose to increase the voltage in order to combat the cold conditions of space that are not favorable to the cutdown method of heating. Electrolytic capacitors lose voltage in cold conditions. The max voltage of the capacitor bank is calculated below.

$$V_{T} = 3V_{C}$$

The voltage of the capacitor bank is 8.1V.

Putting capacitors in series reduces their capacitance. Thus, we added three capacitors in parallel to the original three capacitors in order to increase the capacitance. Thus, more charge could be stored. The capacitance of the capacitor bank (C1-C6) is calculated below.

$$C_{T} = \frac{1}{\frac{1}{C_{1}} + \frac{1}{C_{3}} + \frac{1}{C_{5}}}+\frac{1}{\frac{1}{C_{2}} + \frac{1}{C_{4}} + \frac{1}{C_{6}}}$$
The capacitance of the capacitor bank is 16.7C. The total charge stored in this capacitor bank if it were charged at 8.1V is calculated below.

$$V=\frac{Q}{C}$$

The total charge stored is 135C. This is significantly more than the charge needed for the cutdown to work (4.7C). Although excessive, this ensures that the cutdown will have enough charge to function.

## Transistor Relay System

The cutdown circuit has two power sources. VCC is the constant voltage source supplied by the arduino. V--> is the cutdown signal outputted from the arduino. Both voltage sources are 5V. Q1 is a NPN transistor. NPN transistors work by letting current flow from the collector to the emitter when there is a voltage applied at the base. We used a TIP31C NPN transistor. The TIP31C requires between 1.8V and 5V of current at the base to go on. We used a resistor between the arduino and transistor in order to draw enough current for the base of the transistor. When the transistor turns on, the arduinos constant 5 volt source can send current through the switching section of a relay (K1). When the relay is turned on, the capacitor bank (C1 - C6) can discharge through the nichrome wire (R2).

There are two reasons for using of both a transistor and a relay.  First, transistors tend to burn out more readily when their maximum collector current is exceeded than when a relay’s maximum carrying current is exceeded. Both the relay we used and the transistor have a maximum current allowance of 3A. The discharging of the capacitor bank releases a very high current. The discharge section of the circuit would benefit from running through a component that could handle more current before breaking. This is why a relay was used for the cutdown discharge and not a transistor. Additionally, the relay needs to collect more current to go on than the transistor does. Arduino signals cannot supply the needed current to power the magnetic coils of the relay. Thus, a transistor was used.

## Circuit Walkthrough and Physical Design

First, the V--> sends a signal that goes over R1 and into the base of Q1. Q1 is now on and current can flow from Vcc, through the K1 collector, through the collector and emitter of Q1, and to GND. The relay is now on. C1-6 can discharge over R2. R2 becomes hot and cuts the circuit.

The circuit was soldered on a lightweight protoboard. Electrical tape was used to wrap the entire mechanism. This way, the protoboard would not come in contact with other components in the payload and cause a short circuit. Refer to appendix for diagrams of the cutdown mechanism design and more physical specifications.

## Cutdown Testing

Many tests were conducted where power supplies simulated the voltage outputs of an arduino. Over five tests conducted in this manner were shown to cut 1.4 kg weights down at room temperature. The cutdown was left at 38\degree C for 1 hour and 9 minutes. The cutdown still worked in these conditions.

Next, the cutdown was tested during the tethered launch. A helium balloon was attached at the top of the payload. The cutdown went off while the payload was still on the ground. Pressure sensor fluctuations were so severe that they managed to drop below the specified cutdown pressure. To correct for this, the arduino now tests for the average of 10 readings dropping below the cutdown pressure and not single measurements. Thus, fluctuations will be averaged out and the pressure sensor will not incorrectly trigger the cutdown.

The cutdown was tested on a seperate arduino in a vacuum chamber. The code used for this test follows the averaging method described above and can be found in the appendix. The cutdown successfully burnt through a string to release a 1.4 kg weight when the vacuum chamber dropped below 1.08 kPa. 1.08 kPa is the cutdown pressure specified in the code. This value was chosen because it was the pressure at 100,000 ft.

## Final Code

The pressure testing cutdown code was implement into the main arduino’s code. The cutdown trigger relies on more than just the pressure in the final code. As safety mechanisms to ensure the cutdown doesn't go off too early, the code tests for temperature and time also. The balloon is expected to take longer than 1.5 hours before reaching 100,000 ft. Therefore, the cutdown will not trigger until 1.5 hours have passed. Additionally, the temperature of the atmosphere is less than -10\degree C at 100,000 ft. Thus, the code will not let the cutdown trigger unless the outside temperature readings are less than -10\degree C. The code ensures that the pressure average reading, the temperature reading and the time all fall in the cutdown parameters for 10 readings before triggering the cutdown. This prevents from faulty cutdown triggers due to sensor fluctuations.

There are limitations to the code that prevents the cutdown from triggering too early. If the timer or pressure and temperature sensors fail, the cutdown might never go off. While this error was not encountered because the balloon popped too early (see section 8), this was a possibility.

@@

## Background

A circuit is a set of electronic components that are connected by wires that conduct electricity. Electricity moves through a circuit in the form of charge, and the measurement of the amount of charge (in coulombs) flowing through a circuit per second is called an ampere, or amp. A power source provides a potential energy expressed in volts, which is a relative measure of the electric potential difference across two nodes. These nodes are often expressed as  (+), with voltage, and (-), which is generally at 0 volts. The node with zero volts is often called the ground (GND) because in a power supply it is often connected to the earth, providing a reference point which can be considered at 0V. Electricity wants to move to where there is less voltage, so when these two sides are connected by conductive wires, the current will flow from the positive end to the negative end. When components are connected in series with this flow, they will receive current.  The potential difference expressed in Volts is dissipated in a circuit through a load that takes power. In simple circuits, this can be considered a resistor. The Voltage potential dissipated by a load with resistance $r$ can be related to the current across the resistance with ohm’s law: $V=IR$. The power law states that the power dissipated by an electronic component in Watts can be found with $P=IV$, so therefore a component connected in series with a circuit with current $I$ dissipated through resistance $r$ is $P=I^2R$.

A simple circuit is one where an LED is connected in the middle of the current, which will light up the LED. However, the full voltage over an LED with low resistance would provide too much current for the LED to handle, so the voltage needs to be lowered, which is what a resistor does. Resistors reduce current flow by providing electrical resistance and dissipating power and therefore voltage in the form of heat and are used to control the amount of voltage that reach electrical components.

```{r simple_circuit, fig.cap="\\label{fig:figs} A simple example circuit"}
include_graphics("assets/simple_circuit.png")
```

Arduino is a software and hardware company that produces a variety of microcontrollers called Arduinos that can be programmed. The Arduinos can interact with buttons, lights, motors, the internet, and much more. The microcontrollers have a built-in clock, memory, and voltage regulator. The arduino operates based on the code that is uploaded to it through the Arduino software environment, which provides an development environment to program the arduino using C++ with special arduino libraries built in, and a custom compiler. The way that the arduino works with different components is through pins. These pins can serve either to read input or as output. The first type of pins are those that deal with power, with the Arduino having both pins that output voltage (+) and that ground the circuit (-). Another type of pins are analog pins, which read in a signal from an analog sensor and convert that into a digital value. Conversely, digital pins can read input from digital sensors, but can also output power like to an LED. There are multiple ways to communicate with the Arduino. One way is Serial, which uses universal asynchronous receiver/transmitter (UART) circuitry. Serial UART is asynchronous and does not use the built in Arduino clock for data transfer, which reduces the number of wires and pins needed to communicate. This is how the arduino communicates with the computer programming it, for example. Another method of communication is called $I^2C$ communication which uses two pins on the arduino called SDA and SCL. $I^2C$ communication allows for multiple “slave” chips to communication with multiple “masters” which in our case are Arduinos.

## Electrical Design

The circuit consists of two Arduinos, one Mega ATK and one Uno. The Mega ATK is our primary microcontroller and controls the logger, transmitter, and receives input from every sensor except for the accelerometer. The accelerometer and another microSD logger are the only two components on the secondary Arduino Uno. The reason for the accelerometer needing it’s own Arduino is because the $I^2C$ library it required was not compatible with the other devices running on $I^2C$, such as the BME280 and spectrometer. The primary Arduino is powered by three batteries, and one battery powers the secondary arduino and the methane sensor. The methane sensor uses a large amount of power so it is with the secondary Arduino’s power supply so that if the sensor uses too much power it will not disrupt the majority of the data collection and sensors. All batteries go through 1 Amp 5 Volt voltage regulators, one for the primary Arduino and one for the secondary Arduino and methane sensor. The built in voltage regulator on the Arduino was not powerful enough and causing the sensors to malfunction so exterior regulators were included. Every sensor connects individually to either an analog pin or digital pin, or is one of the three sensors that connect to the SDA and SCL pins.

## Software Design

```{r code_flow_chart, out.height="400px", fig.align='center', fig.cap="\\label{fig:figs} Code Flow Chart"}
include_graphics("assets/flow_code.jpg")
```
